---
title: "assignment.Rmd"
author: "Ieva Gudelytė"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub) # to fetch genomes, annotations, etc.
  library(ensembldb)     # to manipulate ensembl gene annotations
  library(GenomicRanges) # data structure for genomic ranges 
  library(epiwraps)      # visualization
  library(Rsubread)      # read alignment
  library(rtracklayer)   # read/write genomic file formats
  library(Biostrings)    # handling sequences
  library(Rfastp)        # read QC/trimming
  library(MACSr)
})
ah <- AnnotationHub()
```

# Downloading the raw reads (Drosophila ChIP-seq for the protein CTCF)

```{r, eval = FALSE}
options(timeout = 3600)

download.file("https://www.encodeproject.org/files/ENCFF127RRR/@@download/ENCFF127RRR.fastq.gz", "ctcf_chipseq.fastq.gz", mode = "wb")
```

# QC and trimming

```{r}
dir.create("rfastp.trimmed")

do_qc <- function(x) {
  Rfastp::rfastp(x, thread = 4, overrepresentationAnalysis = TRUE,
                 outputFastq = file.path("rfastp.trimmed/", gsub("\\.fastq\\.gz$", "", basename(x))))
}

qc <- do_qc("ctcf_chipseq.fastq.gz")
```

```{r}
Rfastp::curvePlot(qc, curve = "quality_curves")
```

# Alignment

## Building a genome index for mapping

```{r, eval=FALSE}
# we get the genome sequence from AnnotationHub
genome <- ah[["AH106160"]]
# we create a new directory that will contain the genome index
dir.create("BDGP6_genome")
# we write the genome sequence in fasta format
export(import.2bit(genome), "BDGP6_genome/genome.fasta.gz", compress=TRUE)
# we build a Rsubread index
Rsubread::buildindex("BDGP6_genome/rsubread", reference="BDGP6_genome/genome.fasta.gz")
```

## Alignment

```{r}
dir.create("aligned")
align.stats <- Rsubread::align(index = "BDGP6_genome/rsubread", type = "dna",
                               readfile1 = "rfastp.trimmed/ctcf_chipseq_R1.fastq.gz",
                               output_file = "aligned/ctcf.bam",
                               nthreads = 6, sortReadsByCoordinates = TRUE)
align.stats
```

3505377 reads (91.8 %) were mapped.

# Peak calling

(I had a lot of trouble making this work, is there a better solution?)

```{r}
system('wsl bash -lc "cd \"/mnt/c/Users/Ieva/Desktop/Išrūšiuoti/ETH/Academics/sem2/Bioinformatic_approaches_to_regulatory_genomics_and_epigenomics/week03\" && mkdir -p peaks && ~/macs3env/bin/macs3 callpeak --outdir peaks -n ctcf --gsize dm -t aligned/ctcf.bam"')
```

```{r}
peaks <- rtracklayer::import("peaks/ctcf_peaks.narrowPeak")
peaks
```
5405 peaks were found.

## Finding overlaps betwen peaks and genes

```{r}
query_results <- query(ah, c("Drosophila melanogaster", "EnsDb"))
query_results
query_results$rdatadateadded
```

```{r}
ensdb <- ah[["AH119285"]]
ensdb
```

```{r}
fly_genes <- genes(ensdb)
fly_genes
```

```{r}
o <- findOverlaps(peaks, fly_genes)
head(o)

overlapping_peaks = peaks
overlapping_peaks$overlappingGenes <- NA
overlapGenesPerRegion <- splitAsList(fly_genes$gene_name[subjectHits(o)], queryHits(o))
overlapping_peaks$overlappingGenes[as.integer(names(overlapGenesPerRegion))] <- overlapGenesPerRegion

overlapping_peaks
```

## Plotting the signal around one of the peaks that is located inside a gene:

```{r}
plotSignalTracks(list(coverage = "aligned/ctcf.bam", peaks = peaks), region = peaks[410],
                 extend = 500, tracks.params = list(ylim = c(0, 70)))
```
