---
title: "assignment.Rmd"
author: "Ieva Gudelytė"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub) # to fetch genomes, annotations, etc.
  library(ensembldb)     # to manipulate ensembl gene annotations
  library(GenomicRanges) # data structure for genomic ranges 
  library(epiwraps)      # visualization
})
```

Using AnnotationHub, find and download the following annotations data:
The mouse (Mus Musculus) EnsDb object, version 102, genome build GRCm38

```{r}
ah <- AnnotationHub()
query_results <- query(ah, c("Mus musculus", "EnsDb", "102", "GRCm38"))
query_results
```

```{r}
annotations1 <- ah[["AH89211"]]
annotations1
```

The mouse genome sequence ( dna_sm ) in TwoBit/2bit format for GRCm38

```{r}
query_results <- query(ah, c("Mus musculus", "dna_sm", "GRCm38", "TwoBit"))
query_results
query_results$rdatadateadded
rev(query_results$ah_id)[1]
```

```{r}
annotations2 <- ah[["AH88477"]]
annotations2
```

The drosophila melanogaster genome sequence ( dna_sm ) in TwoBit/2bit format for BDGP6

```{r}
query_results <- query(ah, c("Drosophila melanogaster", "dna_sm", "BDGP6", "TwoBit"))
query_results
query_results$rdatadateadded
rev(query_results$ah_id)[1]
```

```{r}
annotations3 <- ah[["AH106160"]]
annotations3
```

Using the mouse EnsDb, find the following:
How many different ensembl gene IDs and gene symbols are there for protein-coding genes?

```{r}
mouse_genes <- genes(annotations1)
mouse_pcg <- mouse_genes[mouse_genes$gene_biotype=="protein_coding"]
mouse_pcg
length(unique(mouse_pcg$gene_id))
length(unique(mouse_pcg$gene_name))
```
From the output we see that there are 22287 ensembl gene IDs and 21964 gene symbols for protein-coding genes.

Plot the distribution of the (spliced) length of protein-coding transcripts
(tip: this will require you to extract exons of protein-coding transcripts from the database, and split them by transcript, before summing the width of the exons of each transcript)

```{r}
mouse_transcripts <- transcripts(annotations1)
mouse_pct <- mouse_transcripts[mouse_transcripts$tx_biotype=="protein_coding"]
mouse_exons_by_tx <- exonsBy(annotations1, filter = TxIdFilter(mouse_pct$tx_id))
mouse_exons_by_tx
```

```{r}
mouse_exon_widths <- width(mouse_exons_by_tx) 
mouse_exon_widths
mouse_exon_width_sum <- sum(mouse_exon_widths)
```

```{r, echo=FALSE}
hist(mouse_exon_width_sum, main = "Histogram of the (spliced) length of protein-coding transcripts", xlab = "Length (bp)", xlim = c(0, 20000), ylim = c(0, 25000), breaks = 100)
```