---
title: "assignment"
author: "Ieva Gudelytė"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(GenomeInfoDb) # for matching chromosome naming styles
  library(SummarizedExperiment)
  library(edgeR) # differential gene expression using neg.bin stats
  library(sechm) # for plotting (features*samples) heatmaps
  library(memes) # for the meme-based methods
  library(motifmatchr) # for scanning sequences for matches of given motifs
  library(Biostrings) # for handling sequences
  library(MotifDb) # database of motifs
  library(TFBSTools) # for handling some motif formats
  library(universalmotif) # for converting motifs to various formats
  library(PWMEnrich) # for R-based motif enrichment analysis
  library(AnnotationHub)
})
ah <- AnnotationHub()
```

# Downloading the data
Downloading TF REST peak counts in liver tissue of male adult Homo sapiens with non-obstructive coronary artery disease.

```{r, eval=FALSE}
download.file("https://www.encodeproject.org/files/ENCFF044GNV/@@download/ENCFF044GNV.bed.gz", "REST_peaks.bed.gz")
```

```{r}
peaks <- rtracklayer::import("REST_peaks.bed.gz", format="NarrowPeak")
seqlevelsStyle(peaks) <- "Ensembl"  # to change the convention of the chromosome names to ensembl
```

Downloading the GRCh38 genome assembly:

```{r}
query_results <- AnnotationHub::query(ah, c("Homo sapiens", "GRCh38", "dna_sm", "TwoBit"))
query_results
```

```{r}
genome <- ah[["AH106285"]]
genome_seqs <- import(genome)
```

# Motif scanning

Motif scanning aims at finding the occurrences of a known motif (i.e motif 'matches') in a set of sequences.

## Getting the desired motif

```{r}
# we search for "REST" in the motif database
motifs <- MotifDb::query(MotifDb, c("REST"))
# there are several matching motifs:
names(motifs)
```

```{r}
# we select one motif:
motif <- motifs[["Hsapiens-HOCOMOCOv10-REST_HUMAN.H10MO.A"]]
motif
# we visualize it:
view_motifs(motifs[1])
```

## Using motifmatchr:

```{r}
# we need to convert the motif to a format that this package will accept
motif_formatted <- convert_motifs(motif, class="TFBSTools-PWMatrix")
moi <- motifmatchr::matchMotifs(motif_formatted, subject=peaks, genome=genome_seqs, out="positions") # we specify that the function should return the genomic positions of each motif match
# we scanned for just one motif, so we get the results for that motif
moi <- moi[[1]]
head(moi)
hist(moi$score)
```

```{r}
# of all the peaks, what proportion contains a motif for the factor?
table(overlapsAny(peaks, moi))
print(paste0("Total peaks: ", length(peaks)))
```

Of the 36521 peaks, 5774 (15.81%) contain a motif

## Scanning the whole genome

To scan the whole genome, one needs to use the genome sequence as input subject.

```{r}
# run the scan:
motif_across_genome <- matchMotifs(motif_formatted, subject=genome_seqs, out="positions")[[1]]
# when inputing matchMotifs directly with sequences, one has to do the following 
# to transform the output to GRanges:
names(motif_across_genome) <- names(genome_seqs)
motif_across_genome <- as(motif_across_genome, "GRanges")
head(motif_across_genome)
```

```{r}
# of all instances of the motif in the genome, what proportion is bound by the factor (i.e. has a peak)?
table(overlapsAny(motif_across_genome, peaks, type="within"))
print(paste0("Total instances: ", length(motif_across_genome)))
```

Of the 567090 motif instances, 6900 (1.22%) overlap a peak